A Plasma Torus Around a Young Low-Mass Star

### 2025-03-10 paper/ms.tex
update

\begin{abstract}
\normalfont
% v1: removed a sentence for wordcount.  v0 is under abstract_title.txt
Approximately one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting clumps of opaque material corotating with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
dark starspots or bright faculae distributed across the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a young ($t$=16\,Myr), rapidly-rotating
($P$=3.9\,hr) CPV, TIC~141146667. The spectra show coherent
sinusoidal Balmer emission at up to four times the star's equatorial
velocity, demonstrating the presence of extended clumps of
circumstellar plasma --- a plasma torus.  Given that long-lived
condensations of cool ($10^4$ K) plasma can persist in the hot
($10^6$ K) coronae of stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that such condensations can become
optically thick around the lowest-mass stars, although the exact
source of opacity remains unclear.
\end{abstract}

@@

%\subsection{Introduction}
M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with water on their surfaces
\cite{NAP26141}.  Investment with JWST has proceeded accordingly.  How
an M dwarf's evolution influences its planets---especially the
retention of their atmospheres--- has concurrently become a major
theme in exoplanet and stellar astrophysics.  Previous work has
established that most M dwarfs host close-in planets
\cite{Dressing2015} that are often subject to long
circumstellar disk lifetimes \cite{Ribas2015}, to large doses of UV
radiation \cite{France2013}, and to a high incidence of flares and
coronal mass ejections \cite{Gunther2020}.  Nevertheless, despite excellent
work in these areas, the properties of the circumstellar plasma and
magnetospheric environments that bathe young, close-in exoplanets have
remained challenging to quantify.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  Figure~\ref{fig:lc} highlights the main object of
interest in this article, but over one hundred analogous objects have
now been found by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
defined by their highly structured and periodic optical light curves, 
and most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \cite{Koen2021}.  Currently, the main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features.  However, no independent evidence has
yet been acquired for the presence of circumstellar material around these
objects.  Since transiting circumstellar clumps would geometrically
imply an intrinsic occurrence a few to ten times the observed rate,
the question of whether circumstellar clumps exist in these systems
has the potential to be applicable to 10-30\% of M dwarfs during their
early lives.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young BAFGKM stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy has
shown periodic high-velocity absorption and emission in Balmer lines
such as H$\alpha$, caused by long-lived, corotating clumps of cool
plasma \cite{CollierCameron1989,Donati2000,Dunstone2006,Skelly2008}.
Such clumps are forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
magnetic field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.
However, none of these spectroscopic variables have shown any
photometric anomalies \cite{Bouma2024}, leaving open the issue of
whether they are related to CPVs.
A circumstantial argument that they might is that CPVs do respond to
sudden magnetic field changes: the otherwise long-lived eclipse
features often disappear following stellar flares
\cite{Stauffer2017,Bouma2024}.

In this study, we present the first observations of corotating clumps
of cool plasma around a CPV.  We identified TIC~141146667 in previous
work \cite{Bouma2024} by searching the TESS two-minute data for stars
showing periodic variability with at least three sharp dips per cycle.
We selected the star for spectroscopic observations because its
brightness and rapid rotation offered sensitivity to small variations
in the line profiles.  We observed it for five hours on UT 2024-02-17
using the High Resolution Echelle Spectrometer (HIRES;
\cite{vogt_hires_1994}) on the Keck I 10m telescope.  TESS observed
the star from UT 2024-02-05 to UT 2024-02-26 with a duty cycle of
XX\%.  TESS was finishing a data downlink during the spectral
observations, and photometric data collection resumed three rotation
cycles (12 hours) after the spectra were acquired.  Extended Data
Figure~\ref{fig:fulllc} shows the detailed photometric behavior of the
star before and after the exact epoch of observation; the star
remained sufficiently stable to not affect the interpretation that
follows.

@@

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  Similar to
other CPVs \cite{Bouma2024}, the photometric shape of
TIC~141146667 evolved following the 2022 discovery
data, while nonetheless remaining complex.  In February 2024, the
average photometric signal showed a gradual brightening over 45\% of
the period, followed by a complex eclipse-like feature spanning 55\%
of the period.  This eclipse feature shows two to three
local photometric minima, and one to two local maxima.
Its W-shape is suggestive of eclipse geometries seen in forward models
of warped plasma tori \cite{Townsend2008}.

@@

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to be corotating with the star.  The
velocity semi-amplitude of the sinusoids gives the distance
of these clumps from the stellar surface: 2.1-2.7\,$R_\star$ for the
closer clump, and 3.9\,$R_\star$ for the other.  This material's
motion, rather than being Keplerian, can only be explained by plasma
being dragged along with the rotating stellar magnetic field.  These
clumps transit in front of the star when passing from negative to
positive velocity.

% * keck/hires: behavior at v/veq<1
The behavior within the stellar H$\alpha$ line core, at $|\Delta v /
v_{\rm eq}|<1$, is more complex than outside it.  For stars of
this age and spectral type, one would expect emission in the line core
to be generated in the stellar chromosphere
and then modulated by any occulting material capable of absorbing or
emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the behavior from
$\phi$=0.4-1.2 has a simple interpretation: from $\phi$=0.4-0.9, a hot
region first gradually crosses the stellar line profile, followed from
$\phi$=0.7-1.2 by the transit of a cool region.  Phases $\phi$$<$0.4
show a mix of similar events, though the time sampling is
sufficiently coarse that the interpretation is less clear.  A final
exercise to quantify the behavior in the line core is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux are usually correlated with the
broadband variability, except at $\phi$=0.5, during the transit of the
higher-velocity clump and the occultation of the lower-velocity clump.

@@

Spectra of magetically-active, rapidly rotating stars with a wide range of masses
have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
sharp transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to 
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.  The usual interpretation for
such variability comes from a loose analogy to
quiescent solar prominences, which are cool
condensations of plasma in the solar corona that can last days to
weeks \cite{VialEngvold2015}.
In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic or centrifugal force
capable of sustaining them.  However for stars with magnetospheric
radii $R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by a plasma parcel can have a local minimum
outside $R_{\rm c}$, enabling the material to be sustained for
much longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.

Our Keck/HIRES observations are the first reported time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present in at least one such star.
Characteristic densities and masses of these
clumps are $n \sim 10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see
Supplementary Methods Section~\ref{subsec:model}), a similar density
to solar prominences, but ten to one hundred times more massive.  This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} since such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios
in which the circumstellar material is made only of dust are also
ruled out.  While dust may be present, to explain the H$\alpha$
emission the circumstellar clumps must include plasma with a
significant population of hydrogen atoms in the $n$=3 excited state.
While this plasma is undoubtedly sculpted by the star's magnetic
field, it could plausibly originate from three sites: the star, an old
and undetected disk, or outgassing rocky bodies.  This latter
possibility would render CPVs as extrasolar analogs of the Jupiter-Io
plasma torus (CITE).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars that with radiatively-driven winds
that can accumulate into
warped plasma tori \cite{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma sculpted by
tilted-dipole magnetic fields, and the transits of
these clumps are thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  For $\sigma$~Ori~E and almost
all of its analogs, the result is light curves that appear ``simple'',
resembling those of eclipsing binaries \cite{Townsend2008}.  The two known
exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: first,
the complexity of CPVs may be a direct consequence of magnetic fields
with highly multipolar contributions.  Second, CPVs could be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young
pre-main-sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
(magneto)-hydrodynamic modeling to go beyond previous work
\cite{Townsend2005,Townsend2008} and to explore the effects of
non-dipolar field contributions.
In particular, dynamo simulations of fully-convective M dwarfs have
suggested the possibility that global-scale mean fields can be
confined to a single hemisphere \cite{Brown2020}; such fields would
yield accumulations of circumstellar material quite different
from those that have previously been explored \cite{Townsend2008}.
Observationally, spectropolarimetry has the potential to
assess both the field strength and topology.  An independent
probe could also be connected the recent work \cite{Kaur2024} showing
that CPVs are variable radio emitters with emission components
that can be both persistent, and also short-lived and highly
polarized.  In particular, detecting radio emission
from the electron cyclotron maser instability would
provide a direct measurement of the field strength at the site of the
emitting region.

It is currently unclear what, if any, relationship CPVs have to the
close-in exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the phenomenon
occurs when clumps of circumstellar material transit the star.  The
implied geometric correction suggests that an appreciable minority
(3-30\%) of M dwarfs -- the rapidly rotating ones with centrifugal
magnetospheres -- have similar circumstellar environments to the CPVs.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the epoch of
  spectroscopic observation (green bar). 
  {\bf Panel a}: TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  {\bf Panels b,c}: Folded light curve before and after spectroscopy.
  While some evolution in the light curve morphology occurred between
  the two epochs, the large, complex eclipse feature remained present.
  }
  \label{fig:fulllc}
\end{figure}

@@

Figure~\ref{fig:fulllc} shows the data from Sector~75, which were
acquired using the camera that points closest to the ecliptic.  The
largest gap in coverage is from BTJD 3356.0~-~3358.5, and includes
the Keck/HIRES observation epoch (green bar).   There are no flux
measurements reported during this interval because the Earth was
within 25$^\circ$ the camera's boresight, yielding high levels of
scattered light.  From BTJD 3359.4~-~3362.0, the Moon then passed
within 25$^\circ$ of the camera's boresight.  Based on the observed
level of scattered light in the optimal TIC~141146667 aperture, we
manually masked out times from 3359.4~-~3360.13, and judged the
remainder of the data during the lunar approach to be usable.  The
small data gaps from BTJD 3353.55~-~3353.77 and from BTJD
3360.12~-~3360.33 were caused by data downlinks at the spacecraft's
perigee and apogee, respectively.

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours over a
second-half night spanning UT 2024-02-17 10:47 to UT 2024-02-17
16:13.  The star's airmass over this window spanned $z$=1.2-2.2, and
we opted for a fixed 15 minute cadence over the entire sequence,
except for a final 10 minute exposure due to increasing sky
brightness at sunrise.  We observed without the iodine cell and used
the C2 decker (0$\farcs$86$\times$14$\farcs$0) in the red instrument
configuration, yielding a spectral resolution $R$$\approx$45{,}0000
($\delta v$$\approx$6.7\,\kms; $\delta v / v_{\rm
eq}$$\approx$0.05).  We binned the CCD readout by a factor of three
in the spatial dimension, yielding $\approx$1,000 photons (S/N=33)
per pixel in the continuum at 6500\,\AA, at minimum airmass.  Strong
winds contributed to 1\farcs2$\pm$0\farcs2 seeing over the night,
but conditions were otherwise favorable.  We reduced the
echelleogram to a one-dimensional spectrum using the standard
techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec} shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

{\it Radial Velocity}---We measured radial velocities of
TIC~141146667 from each of our spectra using a pipeline that we
developed specifically for rapidly rotating stars.
Our method is based on template-matching against synthetic spectra
produced by the PHOENIX stellar atmosphere code \cite{Husser2013}.
We used the PHOENIX models with solar metallicity and alpha element
abundances, and calibrated our absolute velocity zero-points as well
as viable orders for velocity measurement using the
standard stars described by \cite{Chubak2012}.
We used velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.
We used \texttt{barycorrpy} \cite{Kanodia2018} to calculate the
velocity corrections due to Earth's motion around the solar system
barycenter and due to Earth's daily rotation about its axis.
Our analysis pipeline reproduces the systemic velocities reported by
\cite{Chubak2012} for their velocity standards with an RMS of
0.66\,\kms.

For TIC~141146667, we measured the velocities from each of our HIRES
spectra using regions near the K~I (7700\,\AA) resonance line and
three TiO bandheads (5160\,\AA, 5450\,\AA, and 5600\,\AA).  These
regions were selected because they provided the best matches between
the synthetic and observed spectra.  The resulting redshift
measurements were averaged together.  The star's 130$\pm$4\,\kms\
equatorial velocity and near-edge on viewing orientation yielded a
line broadening $\Delta \lambda$$\approx$3\,\AA.  We used the scatter
of resulting velocity measurements between orders to assign the RV
uncertainty at each epoch.  The uncertainty-weighted mean systemic
velocity over all epochs on UT 2024-02-17 was
$\gamma$=0.6$\pm$1.5\,\kms.  The relative radial velocities about this
mean are given in Table~\ref{tab:rv}.
% from calc_max_amplitude_sinusoid_rv_timeseries.py
% from K_to_msini.py, assuming Mstar = 0.18

>>>>

\begin{abstract}
\normalfont
Approximately one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material corotating with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
starspots or faculae distributed across the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a young ($t$=16\,Myr), rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the presence of extended clumps of circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
material and its microphysical opacity remain unclear.
\end{abstract}

@@

%\subsection{Introduction}
M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water
\cite{NAP26141}.  Investment with JWST has proceeded accordingly.  How
an M dwarf's evolution influences its planets---especially the
retention of their atmospheres--- has concurrently become a major
theme in exoplanet and stellar astrophysics.  Previous work has
established that most M dwarfs host close-in planets
\cite{Dressing2015} that are often subject to long
circumstellar disk lifetimes \cite{Ribas2015}, to large doses of UV
radiation \cite{France2013}, and to a high incidence of flares and
coronal mass ejections \cite{Gunther2020}.  Nevertheless, despite excellent
work in these areas, the properties of the plasma and magnetospheric
environments that bathe young, close-in exoplanets have remained
challenging to quantify.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  Figure~\ref{fig:lc} highlights the main object of
interest in this article, but over one hundred analogous objects have
now been found by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
phenomenologically identified based on their highly structured and
periodic optical light curves;
most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \cite{Koen2021}.  Currently, the main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features.  However, no independent evidence has
yet been acquired for the presence of circumstellar material around these
objects.  
Transiting circumstellar clumps would geometrically imply an intrinsic
occurrence a few to ten times the observed rate.  The question of
whether circumstellar clumps exist in these systems could therefore be
relevnat for 10-30\% of M dwarfs during their early lives.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy for
stars with a wide range of masses has shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,Donati2000,Dunstone2006,Skelly2008}.
Such clumps are forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.
To date, none of these spectroscopic variables have shown any
photometric anomalies \cite{Bouma2024}, leaving open the issue of
whether they are related to CPVs.
%A circumstantial argument that they might is that CPVs do respond to
%sudden magnetic field changes: the otherwise long-lived eclipse
%features often disappear following stellar flares
%\cite{Stauffer2017,Bouma2024}.

In this study, we present the first observations of corotating clumps
of cool plasma around a CPV.  We identified TIC~141146667 in previous
work \cite{Bouma2024} by searching the TESS two-minute data for stars
showing periodic variability with at least three sharp dips per cycle.
We selected the star for spectroscopic observations because its
brightness and rapid rotation offered sensitivity to small variations
in the line profiles.  We observed for five hours on UT 2024-02-17
using the High Resolution Echelle Spectrometer (HIRES;
\cite{vogt_hires_1994}) on the Keck I 10m telescope.  TESS observed
the star from UT 2024-02-05 to UT 2024-02-26 with a duty cycle of
XX\%.  
At the exact instant of spectral observations, TESS was performing a
data downlink; photometric data collection resumed three rotation
cycles (12 hours) after the spectra were acquired.  Extended Data
Figure~\ref{fig:fulllc} shows the detailed photometric behavior of the
star before and after the exact epoch of observation; the photometric
morphology is similar before and after the data gap.

@@

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  Over
timescales of years, CPVs maintain the same period, while their
photometric morphology evolves.  TIC~141146667 indeed evolved relative
to the February 2022 discovery data \cite{Bouma2024}.  In the February
2024 data, the average photometric signal showed a small brightening
over 45\% of the period, followed by a complex flux dip spanning 55\%
of the period.  This eclipse feature shows two to three local
photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in some forward models of warped
plasma tori \cite{Townsend2008}.

@@

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to be corotating with the star.  
Based on its period, this material's motion is not Keplerian; it can
only be explained by plasma being dragged along with the rotating
stellar magnetic field. 
The velocity semi-amplitude of the sinusoids then gives the distance
of the clumps from the stellar surface: 2.1-2.7\,$R_\star$ for the
closer clump, and 3.9\,$R_\star$ for the other.   These clumps transit
in front of the star when passing from negative to positive velocity.
This implies that the spectroscopic transit of the 2.1-2.7\,$R_\star$
clump occurs simultaneously with the sharp photometric spike visible
in the TESS data.

% * keck/hires: behavior at v/veq<1
The behavior within the stellar H$\alpha$ line core, at $|\Delta v /
v_{\rm eq}|<1$, is more complex than outside it.  For stars of
this age and spectral type, one would expect emission in the line core
to be generated in the stellar chromosphere
and then modulated by any occulting material capable of absorbing or
emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the behavior from
$\phi$=0.4-1.2 has a simple interpretation: from $\phi$=0.4-0.9, a hot
region first gradually crosses the stellar line profile, followed from
$\phi$=0.7-1.2 by the transit of a cool region.  Phases $\phi$$<$0.4
show a mix of similar events, though the time sampling is
sufficiently coarse that the interpretation is less clear.  A final
exercise to quantify the behavior in the line core is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux are usually correlated with the
broadband variability, except at $\phi$=0.5, during the transit of the
3.9\,$R_\star$ clump and the occultation of the lower-velocity clump.

@@

Spectra of magetically-active, rapidly rotating stars with a wide range of masses
have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to 
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.  The usual interpretation for
such spectroscopic variability comes from an analogy to
quiescent solar prominences, which are cool
condensations of plasma in the solar corona that can last days to
weeks \cite{VialEngvold2015}.
In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic or centrifugal force
capable of sustaining them.  However for stars with magnetospheric
radii $R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to be sustained for
much longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.

Our Keck/HIRES observations are the first reported time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present in at least one such star.
Characteristic densities and masses of these
%TODO check; quote N_H instead...
clumps are $n \sim 10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see
Supplementary Methods Section~\ref{subsec:model}), a similar density
to solar prominences, but ten to one hundred times more massive.  This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} since such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios
in which the circumstellar material is made only of dust are also
ruled out.  While dust may be present, to explain the H$\alpha$
emission the circumstellar clumps must include plasma with a
significant population of hydrogen atoms in the $n$=3 excited state.
While this plasma is undoubtedly sculpted by the star's magnetic
field, it could plausibly originate from three sites: the star, an old
and undetected disk, or outgassing rocky bodies.  This latter
possibility would render CPVs as extrasolar analogs of the Jupiter-Io
plasma torus (CITE).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds
that accumulate into
warped plasma tori \cite{Townsend2005,Townsend2008}.  These tori tend
to have dense antipodal accumulations of plasma sculpted by
tilted-dipole magnetic fields, and the transits of
these clumps are thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  For $\sigma$~Ori~E and almost
all of its analogs, the result is light curves that appear ``simple'',
resembling those of eclipsing binaries \cite{Townsend2008}.  The two known
exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: first,
the complexity of CPVs may be a direct consequence of magnetic fields
with highly multipolar contributions.  Second, CPVs could be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young
pre-main-sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The role of the star's magnetic field could be better understood
through new observations, and new theory.  From the theoretical
perspective, there is an urgent need for rigid-field
(magneto)-hydrodynamic modeling to go beyond previous work
\cite{Townsend2005,Townsend2008} and to explore the effects of
non-dipolar field contributions.
In particular, dynamo simulations of fully-convective M dwarfs have
suggested that global-scale mean fields might be
confined to a single hemisphere \cite{Brown2020}; such fields would
yield accumulations of circumstellar material quite different
from those that have previously been explored \cite{Townsend2008}.
Observationally, spectropolarimetry has the potential to
assess both the field strength and topology.  An independent
probe could also be connected the recent work \cite{Kaur2024} showing
that CPVs are variable radio emitters with emission components
that can be both persistent, and also short-lived and highly
polarized.  In particular, detecting radio emission produced by an electron
cyclotron maser instability would provide a measurement of the field
strength at the site of the emitting region.

It is currently unclear what, if any, relationship CPVs have to the
close-in exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the phenomenon
occurs when clumps of circumstellar material transit the star.  The
implied geometric correction suggests that an appreciable minority
(10-30\%) of young M dwarfs -- the rapidly rotating ones with
centrifugal magnetospheres -- host similar circumstellar environments
to the CPVs.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/sf1.pdf}
  \caption{Photometric evolution of TIC 141146667 near the epoch of
  spectroscopic observation (green bar). 
  {\bf Panel a}: TESS simple aperture photometry. 
  The main data gaps were caused by scattered light from the Earth
  (BTJD 3356-3358.5) and Moon (BTJD 3359.5-3360.5).
  {\bf Panels b,c}: Folded light curve before and after spectroscopy.
  Although some evolution in the light curve morphology occurred between
  the two epochs, the large, complex eclipse feature remained present.
  }
  \label{fig:fulllc}
\end{figure}

@@

Figure~\ref{fig:fulllc} shows the data from Sector~75, which were
acquired using the camera that pointed closest to the ecliptic.  The
largest gap in coverage is from BTJD 3356.0~-~3358.5, and includes
the Keck/HIRES observation epoch (green bar).   There are no flux
measurements reported during this interval because the Earth was
within 25$^\circ$ the camera's boresight, yielding high levels of
scattered light.  From BTJD 3359.4~-~3362.0, the Moon then passed
within 25$^\circ$ of the camera's boresight.  Based on the observed
level of scattered light in the optimal TIC~141146667 aperture, we
manually masked out times from 3359.40~-~3360.13, and judged the
remainder of the data during the lunar approach to be usable.  The
small data gaps from BTJD 3353.55~-~3353.77 and from BTJD
3360.12~-~3360.33 were caused by data downlinks at the spacecraft's
perigee and apogee, respectively.

{\it Spectroscopy:}
We observed TIC~141146667 ($V$=16.2) with Keck/HIRES for five hours over a
second-half night spanning UT 2024-02-17 10:47 to UT 2024-02-17
16:13.  The star's airmass over this window spanned $z$=1.2-2.2, and
we opted for a fixed 15 minute cadence over the entire sequence,
except for a final 10 minute exposure due to increasing sky
brightness at sunrise.  We observed without the iodine cell and used
the C2 decker (0$\farcs$86$\times$14$\farcs$0) in the red instrument
configuration, yielding a spectral resolution $R$$\approx$45{,}0000
($\delta v$$\approx$6.7\,\kms; $\delta v / v_{\rm
eq}$$\approx$0.05).  We binned the CCD readout by a factor of three
in the spatial dimension, yielding $\approx$1,000 photons (S/N=33)
per pixel in the continuum at 6500\,\AA, at minimum airmass.  Strong
winds contributed to 1\farcs2$\pm$0\farcs2 seeing over the night,
but conditions were otherwise favorable.  We reduced the
echelleogram to a one-dimensional spectrum using the standard
techniques of the California Planet Survey \cite{Howard2010}.
Figure~\ref{fig:spec}b shows the result in the vicinity of H$\alpha$
without any additional processing.

@@

{\it Radial Velocity}---We measured radial velocities of
TIC~141146667 from each of our spectra using a pipeline that we
developed specifically for rapidly rotating stars.
Our method is based on template-matching against synthetic spectra
produced by the PHOENIX stellar atmosphere code \cite{Husser2013}.
We used the PHOENIX models with solar metallicity and alpha element
abundances, and calibrated our absolute velocity zero-points as well
as viable orders for velocity measurement using the
standard stars described by \cite{Chubak2012}.
We used velocity standards spanning spectral types from G2 to M4 (Barnard's
Star), irrespective of rotation rate.
We used \texttt{barycorrpy} \cite{Kanodia2018} to calculate the
velocity corrections due to Earth's motion around the solar system
barycenter and due to Earth's daily rotation about its axis.
Our analysis pipeline reproduces the systemic velocities of known
velocity standards \cite{Chubak2012} with an RMS of 0.66\,\kms.

For TIC~141146667, we measured the velocities from each of our HIRES
spectra using regions near the K~I (7700\,\AA) resonance line and
three TiO bandheads (5160\,\AA, 5450\,\AA, and 5600\,\AA).  These
regions were selected because they provided the best matches between
the synthetic and observed spectra.  The resulting redshift
measurements over each order were averaged to produce the final measurement.  The star's 130$\pm$4\,\kms\
equatorial velocity and near-edge on viewing orientation yielded a
line broadening $\Delta \lambda$$\approx$3\,\AA.  We used the scatter
of resulting velocity measurements between orders to assign the RV
uncertainty at each epoch.  The uncertainty-weighted mean systemic
velocity over all epochs on UT 2024-02-17 was
$\gamma$=0.6$\pm$1.5\,\kms.  The relative radial velocities about this
mean are given in Table~\ref{tab:rv}.
% from calc_max_amplitude_sinusoid_rv_timeseries.py
% from K_to_msini.py, assuming Mstar = 0.18
